Formalizing the Kolmogorov extension Theorem in Lean

### 2023-09-08 Manuscript/proof_outline.tex
updated including 2.3

\title{Formalizing the Kolmogorov extension Theorem in mathlib4}

\keywords{probability, measure theory, Lean4, formal mathematics, proof assistant, mathlib4}

@@

\begin{abstract}
  We present a formalization of the Kolmogorov extension theorem,
  which is a main building block in construction the (distribution of)
  stochastic processes with arbitrary index sets in Lean4. Our
  approach is based on mathlib, the mathematical library for Lean4. On
  our way, we provide a formalization of the Caratheodory extension
  theorem, which allows to extend a set function on a semi-ring to a
  proper measure on the $\sigma$-algebra generated by the semi-ring. A
  slight generalization of the classical Kolmogorov extension theorem
  allows us to construct the (distribution of) a stochastic process on
  a complete and separable pseudo-metric space (rather than a metric
  space).
\end{abstract}

@@

The goal of our contribution is a formalization of the proof of that theorem in {\tt mathlib4}, the mathematical library of {\tt Lean4} \cite{mathlib2020lean}. The Kolmogorov extension theorem is on the interface between measure theory and probability theory. Here, we rely on a decent amount of formalized mathematics in the measure-theory part of {\tt mathlib4} (outer measures, above all), while not requiring any specific previous formalization of probability theory. (In fact, most of our results are formulated in terms of finite rather than probability measures.)

@@

The main steps in our construction, which were previously missing in {\tt mathlib4}, are (more mathematical details are below, and in standard textbooks on Probability Theory, e.g.\ \cite{Klenke2013, Kallenberg2020}):
\begin{enumerate}
\item a formalization of (the set-system of) semi-rings (see Definition~\ref{def:semi} below); 
\item adding some main lemmas to measure theory, e.g.\ continuity of a measure at the emptyset (see e.g.\ Lemma~\ref{l:halbextRing}, Proposition~\ref{P:stetigmass}, Lemma~\ref{l:stetigcompact} below);
\item a proof that a single probability measure on a Polish space is inner regular with respect to compact sets (see Definition~\ref{def:innerreg}; in fact, we proved a slight generalization using pseudo-metric spaces, found in Lemma~\ref{L:relcoPol} below); 
\item the classical Carathéodory extension theorem, providing us a candidate for the measure which we want to construct (see Theorem~\ref{T:masseind}); 
\item the Kolmogorov extension theorem, as based on the previous steps (See Theorem~\ref{T1}).
\end{enumerate}
The Kolmogorov extension theorem has been previously formalized in Isabelle/HOL \cite{Immler2012}. This formalization only works on Polish spaces (rather than on spaces where every finite measure is inner regular with respect to compact sets, see below), and only in the case where all $\alpha_i$'s are identical. However, type theory as the basis of the formalization, as well as several mathematical concepts (e.g.\ projective families of measures or the continuity of a measure in the emptyset) are the same in both formalisations in these two interactive theorem provers.

\subsubsection*{Possible future work}
\sloppy Let us describe some future projects extending {\tt mathlib4} which become possible by our contribution.

\noindent
{\em Instances of stochastic processes:} An obvious application of Kolmogorov's extension theorem is the construction of basic stochastic processes like Markov chains \cite[Section~11]{Kallenberg2020}, the Poisson process \cite[Section~13]{Kallenberg2020} and Brownian motion \cite[Section~14]{Kallenberg2020}. (Note that other Gaussian processes -- indexed by $[0, \infty)$ or $\mathbb R$ -- might as well be constructed the same way. In addition, fields like the Gaussien free field indexed by $\mathbb R^d$ \cite{sheffield2007gaussian} can also be given.) For these tasks, we would have to define the finite-dimensional distributions (Poisson and normally distributed, respectively), and apply the extension theorem. This task requires the formalization of multi-dimensional Poisson and normal distributions, which -- in textbooks -- is usually done using characteristic functions. Since these are not yet part of {\tt mathlib4}, we postpone this task to the future.

@@

\noindent
{\em Theorem of Ionescu-Tulcea and related results:} While Kolmogorov's extension Theorem gives a result for arbitrary $\iota$, but requires some properties of the family $(\alpha_t)_{t\in\iota}$, the Theorem of Ionescu-Tulcea \cite[Theorem 8.24]{Kallenberg2020} can only deal with countable $\iota$, but has no restrictions on $(\alpha_t)_{t\in\iota}$. The proof of the latter, however, uses transition kernels (and an inductive construction of a content, which can be extended to a measure) rather than projective families. However, continuity of a measure at $\emptyset$ is used as well and a formalization would complement our work. In addition, using the Theorem of Ionescu-Tulcea one can define infinitely many independent random variables (arbitrary $\iota$) on a joint probability space \cite[Theorem 8.24]{Kallenberg2020} without requirements on $(\alpha_t)_{t\in\iota}$. All of these results can be used in order to construct countably many independent random variables, which are frequently used in concrete constructions in probability theory. Examples are (one direction of) the Borel-Cantelli-Lemma \cite[Theorem 4.28]{Kallenberg2020}, Kolmogorov's 0-1-law \cite[Theorem 4.13]{Kallenberg2020}, random walks \cite[Chapter 12]{Kallenberg2020}, branching processes  \cite[Chapter 13]{Kallenberg2020}, and percolation  \cite[Chapter 2.4]{Klenke2013}, to name just a few. While the first two are already formalized in \leanline{mathlib4}, it is only clear due to the Kolmorov extension theorem (or the Theorem of Ionesu-Tulcea) that infinitely many independent random variables can indeed be defined on a joint probability space. In this sense, our contribution is also important for giving more sense to already formalised parts of \leanline{mathlib}.

\section{Formalization of the Daniell-Kolmogorov extension}
We start by stating the exact result. We are going to formulate the main result in a modern fashion, as e.g.\ found in Theorem 2.2 of \cite{rao1971projective}, Theorem 7.7.1 of Volume~2 of \cite{bogachev2007measure}, Theorem 15.26 of \cite{guide2006infinite}, or \cite{border1998expository}. Note that these formulations split general assumptions on the underlying space(s) (e.g.\ a metric property) from the property which is needed in the proof (inner regularity with respect to compact sets). Other -- highly readable -- references such as \cite{Billingsley1995} state the extension theorem only in special cases such as $\alpha_t = \mathbb R$ for all $t$. 
%In addition, there are still new proofs found for the general extension theorem \cite{chin2019new}, but they require some results not yet formalized in mathlib4 (Borel isomorphism of $\mathbb R$ to $\{0,1\}^{\mathbb N}$), so we do not follow this route.

\subsection{Formulating the result}
Before we can state the result, we begin with some mathematical notions. We start off from metric spaces, but quickly introduce projective families of measures. In all definitions, $\alpha$ will be some set (or type, since {\tt Lean4} is a dependently typed language, \leanline{α : Type _}).

@@

\begin{remark}[Metric and topological spaces in {\tt mathlib4}] \mbox{}
  \begin{enumerate}
    \item \sloppy In \leanline{Lean4}, we have the \leanline{class
      PseudoEMetricSpace (α : Type u) extends EDist}, where
      \leanline{EDist} comes with an extended distance \leanline{edist
        : α → α → ENNReal}, and the extension to
      \leanline{PseudoEMetricSpace} consists of the components,
      \leanline{edist_self}, \leanline{edist_comm},
      \leanline{edist_triangle} providing the properties of the
      pseudo-metric, and \leanline{toUniformSpace},
      \leanline{uniformity_edist} defining a uniform space from the
      extended pseudo-metric. Such a space does not come with a
      metric, but with a filter on $\alpha \times \alpha$, which
      describes which points in $\alpha$ are near. For example, the
      diagonal of $\alpha \times \alpha$ is a subset of all sets in
      the uniformity; see \cite{james2013topologies} for details. We
      note that a uniform space with a countably generated uniformity
      filter is pseudometrizable, i.e.\ there exists a
      pseudo-metric-space structure that generates the same
      uniformity; see \leanline{UniformSpace.pseudoMetrizableSpace},
      which formalizes a result stated in
      \cite{melikhov2011metrizable}.
    \item \sloppy For topological spaces, we have \leanline{class
      TopologicalSpace(α : Type u) : Type u}, which comes with
      \leanline{IsOpen : Prop}, \leanline{isOpen_univ : IsOpen
        Set.univ}, \leanline{isOpen_inter} and
      \leanline{isOpen_sUnion}, which are exactly the properties of a
      topological space described above.
    \item Any (extended pseudo-)metric on $\alpha$ defines a topology,
      namely the topology generated by $\mathcal H := \{\{y:
      r(x,y)<\varepsilon\}: x \in \alpha, \varepsilon \in
      (0,\infty)\}$. Actually, in \leanline{mathlib4}, as we have seen
      above, every extended pseudo-metric defines a uniform space, and
      a uniform space is a \leanline{class UniformSpace(α : Type u)
        extends TopologicalSpace , UniformSpace.Core} (where the
      uniformity is defined in \leanline{UniformSpace.Core}), which
      connects the \leanline{uniformity} with the topology using
      \leanline{∀ (s : Set α), IsOpen s ↔ ∀ (x : α), x ∈ s → {p |
          p.fst = x → p.snd ∈ s} ∈ uniformity}. In particular, a
      uniform space defines a topological space, which can be used by
      the typeclass system.
  \end{enumerate}
\end{remark}

@@

\begin{remark}[Measur(abl)e spaces in \leanline{mathlib4}]
  \begin{enumerate}
  \item \sloppy The \leanline{class MeasurableSpace α} is very similar
    to a topological space in \leanline{mathlib4}, since it comes with
    \leanline{MeasurableSet' : Set α → Prop},
    \leanline{measurableSet_empty : MeasurableSet' ∅},
    \leanline{measurableSet_compl : (s : Set α) → MeasurableSet' s →
      MeasurableSet' sᶜ}, and \leanline{measurableSet_iUnion},
    i.e.\ properties (i)-(iii) from Definition~\ref{def:sigma}.1.
  \item We are using outer measures in our construction. In
    \leanline{mathlib4}, this is \leanline{structure
      MeasureTheory.OuterMeasure α} which is a function
    \leanline{measureOf : Set α → ℝ≥0∞} such that \leanline{measureOf
      ∅ = 0} (i.e.\ the empty set has measure~$0$), \leanline{∀ {s₁ s₂
        : Set α}, s₁ ⊆ s₂ → measureOf s₁ ≤ measureOf s₂}
    (i.e.\ monotonicity) and \leanline{∀ (s : ℕ → Set α), measureOf (⋃
      (i : ℕ), s i) ≤ ∑' (i : ℕ), measureOf (s i)} (which we call
    $\sigma$-subadditivity). \\ Note that an outer measure is defined
    on $2^\alpha$, whereas a measure is only defined on a subset (the
    $\sigma$-algebra of measurable sets). More precisely: A
    \leanline{MeasureSpace α} extends a \leanline{MeasurableSpace α}
    by a \leanline{Measure α} (which is usually called
    \leanline{volume}). For the measure, we have: \leanline{structure
      MeasureTheory.Measure α [MeasurableSpace α] extends
      MeasureTheory.OuterMeasure}. Here $\sigma$-additivity of a
    \leanline{Measure α} reads
\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  m_iUnion : ∀ ⦃f : ℕ → Set α⦄,
    (∀ (i : ℕ), MeasurableSet (f i)) →
    Pairwise (Disjoint on f) →
    ↑toOuterMeasure (⋃ (i : ℕ), f i) =
    ∑' (i : ℕ), ↑toOuterMeasure (f i)
\end{minted}    
  \end{enumerate}
\end{remark}

@@

\begin{remark}[Generated $\sigma$-algebra, image measure]
  We will frequently need two basic results: 
  \begin{enumerate}
  \item The intersection of any number of $\sigma$-algebras is again a
    $\sigma$-algebra. For this reason, if $\mathcal H \subseteq
    2^\alpha$, we define the $\sigma$-algebra $\sigma(\mathcal H) :=
    \bigcap_{\mathcal F \supseteq \mathcal H \text{ $\sigma$-algebra}}
    \mathcal F$. This is called the $\sigma$-algebra generated from
    $\mathcal H$; see \leanline{MeasurableSpace.generateFrom}.  \\ In
    particular, if $\mathcal O$ is a topology on $\alpha$, we call
    $\mathcal B := \sigma(\mathcal O)$ the Borel $\sigma$-algebra on
    $\alpha$. In \leanline{mathlib4}, this is
\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  def borel (α : Type u) [TopologicalSpace α] :
    MeasurableSpace α :=
    generateFrom { s : Set α | IsOpen s }
\end{minted}
    In addition, \leanline{mathlib4} provides a similar notion, which is
\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  class OpensMeasurableSpace (α : Type*)
    [TopologicalSpace α] [h : MeasurableSpace α] :
    Prop where borel_le : borel α ≤ h
\end{minted}
    Here, all open sets are measurable, so the $\sigma$-algebra defining
    \leanline{h} might be larger than the Borel $\sigma$-algebra. We will
    need both notions in our proofs.
  \item Let $(\alpha, \mathcal F)$ and $(\beta, \mathcal G)$ be
    measurable spaces. Some $f : \alpha \to \beta$ is called
    measurable (with respect to $\mathcal F$ and $\mathcal G$) if
    $f^{-1} \mathcal G \subseteq \mathcal F$. In \leanline{mathlib4},
    see \leanline{Measurable} and note that this notion usually comes
    with some \leanline{m : Measure α} and $f$ is almost everywhere
    measurable (\leanline{AEMeasurable}) if it coincices with some
    \leanline{Measurable g}, \leanline{m}-almost everywhere.\\ In this
    case, if $\mu$ is a measure on $\mathcal F$, the measure $\nu:
    \mathcal G \to [0, \infty], \nu(B) := \mu (f^{-1}(B))$ is called
    the image (or push-forward) measure of $\mu$ under $f$. We write
    $\nu := f_\ast \mu$; see \leanline{MeasureTheory.Measure.map}.
  \end{enumerate}
\end{remark}

We now come to our contribution, which we will PR to mathlib4: The
extension theorem is on finite measures on product spaces. In
particular, the product is taken over a set of arbitrary
cardinality. The next definition covers the important concept of a
projective family of measures. In short term, we define measures on
any finite subset of indices in a consistent way. Recall that finite
sets (of some type \leanline{ι}) are formalized with \leanline{Finset
  ι}. Such a \leanline{Finset} is defined as a \leanline{Multiset} (a
quotient type of the \leanline{List}-type with respect to permutations
of the list) together with a proof that it contains no duplicates.

@@

\noindent
With this, we can define the projetive family as follows. Note that
\leanline{Lean4} automatically uses the subset relation as
\leanline{[Preorder (Finset ι)]} when \leanline{isProjective} is
called.

@@

Let us remark that (i) \leanline{Lean4} indicates a coercion by
\leanline{↑}, which in this case is from \leanline{Finset} to
\leanline{Set} and (ii) we have \leanline{PairwiseDisjoint (s : Set ι)
  (f : ι → α)} iff the images of any distinct elements of \leanline{ι}
under \leanline{f} are different. (Hence, if \leanline{f = id}, the
usual definition of pairwise disjoint sets unfolds.)

@@

\noindent
An additive $m : \mathcal H \to [0,\infty]$ is formalized as follows:

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  structure AddContent (C : Set (Set α)) where
    toFun : Set α → ℝ≥0∞
    empty' : toFun ∅ = 0
    add' : ∀ (I : Finset (Set α))
      (_h_ss : ↑I ⊆ C) (_h_dis : PairwiseDisjoint
        (I : Set (Set α)) id)
      (_h_mem : ⋃₀ ↑I ∈ C),
      toFun (⋃₀ I) = ∑ u in I, toFun u
\end{minted}

\noindent
In this formalization, we introduced $m(\emptyset)=0$ as a separate
hypothesis with \leanline{empty'}. Note that this also follows
formally from $m(\emptyset) = m(\emptyset \cup \emptyset) = 2
m(\emptyset)$. However, since \leanline{Finset}s are assumed to have
distinct elements, we cannot have \leanline{∅} twice in \leanline{I}.

% extend_content put $\infty$ outside the set system

For the concrete application we have in mind, we need to be more
specific which semiring to use, based on a projective family of
probability measures. We do this using \leanline{kolContent}, which
is based on the following function, we takes a family of measures and
a cylinder set as input:

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  noncomputable def kolmogorovFun
    (P : ∀ J : Finset ι, Measure (∀ j : J, α j))
    (s : Set (∀ i, α i))
    (hs : s ∈ cylinders α) : ℝ≥0∞ :=
    P (cylinders.finset hs) (cylinders.set hs)
\end{minted}

\noindent
With this, we define an \leanline{add_content} on cylinders. (Note
that \leanline{extend_content} in fact defines an additive content on
a semiring.)

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  noncomputable def extendContent
    (hC : SetSemiring C)
    (m : ∀ s : Set α, s ∈ C → ℝ≥0∞)
    (m_empty : m ∅ hC.empty_mem = 0)
    (m_add :
       ∀ (I : Finset (Set α)) (h_ss : ↑I ⊆ C)
       (_h_dis : PairwiseDisjoint (I : Set (Set α)) id)
    (h_mem : ⋃₀ ↑I ∈ C),
    m (⋃₀ I) h_mem = ∑ u : I, m u (h_ss u.prop)) :
    AddContent C

@@

\noindent
From the next lemma, we will need monotonicity and sub-additivity, as
well as $\sigma$-additive $\Rightarrow$ $\sigma$-sub-additive in the
proof of the Carathéodory extension, Theorem~\ref{T:masseind}.

@@

The above definition closely resembles its formalization in
\leanline{mathlib4}:

>>>>

\title{Formalizing the Kolmogorov extension Theorem in Lean}

\keywords{probability, measure theory, Lean, formal mathematics, proof assistant, mathlib}

@@

\begin{abstract}
  We present a formalization of the Kolmogorov extension theorem,
  which is a main building block in construction the (distribution of)
  stochastic processes with arbitrary index sets in Lean. Our
  approach is based on mathlib, the mathematical library for Lean. On
  our way, we provide a formalization of the Caratheodory extension
  theorem, which allows to extend a set function on a semi-ring to a
  proper measure on the $\sigma$-algebra generated by the semi-ring. A
  slight generalization of the classical Kolmogorov extension theorem
  allows us to construct the (distribution of) a stochastic process on
  a complete and separable pseudo-metric space (rather than a metric
  space).
\end{abstract}

@@

The goal of our contribution is a formalization of the proof of that theorem in {\tt Lean}. Here, we are going to use \leanline{Lean4}, since \leanline{mathlib} \cite{mathlib2020lean}, the mathematical library of \leanline{Lean} has recently been ported to \leanline{Lean4}. Subsequently, we aim to make a pull request for the publication of our code in \leanline{mathlib}. The Kolmogorov extension theorem is on the interface between measure theory and probability theory. Here, we rely on a decent amount of formalized mathematics in the measure-theory part of {\tt mathlib} (outer measures, above all), while not requiring any specific previous formalization of probability theory. (In fact, most of our results are formulated in terms of finite rather than probability measures.)

@@

The main steps in our construction, which were previously missing in {\tt mathlib}, are (more mathematical details are below, and in standard textbooks on Probability Theory, e.g.\ \cite{Klenke2013, Kallenberg2020}):
\begin{enumerate}
\item a formalization of (the set-system of) semi-rings (see Definition~\ref{def:semi} below); 
\item adding some main lemmas to measure theory, e.g.\ continuity of a measure at the emptyset (see e.g.\ Lemma~\ref{l:halbextRing}, Proposition~\ref{P:stetigmass}, Lemma~\ref{l:stetigcompact} below);
\item a proof that a single probability measure on a Polish space is inner regular with respect to compact sets (see Definition~\ref{def:innerreg}; in fact, we proved a slight generalization using pseudo-metric spaces, found in Lemma~\ref{L:relcoPol} below); 
\item the classical Carathéodory extension theorem, providing us a candidate for the measure which we want to construct (see Theorem~\ref{T:masseind}); 
\item the Kolmogorov extension theorem, as based on the previous steps (See Theorem~\ref{T1}).
\end{enumerate}
The Kolmogorov extension theorem has been previously formalized in Isabelle/HOL \cite{Immler2012}. This formalization only works on Polish spaces (rather than on spaces where every finite measure is inner regular with respect to compact sets, see below), and only in the case where all $\alpha_i$'s are identical. However, type theory as the basis of the formalization, as well as several mathematical concepts (e.g.\ projective families of measures or the continuity of a measure in the emptyset) are the same in both formalisations in these two interactive theorem provers.

\subsubsection*{Possible future work}
\sloppy Let us describe some future projects extending {\tt mathlib} which become possible by our contribution.

\noindent
{\em Instances of stochastic processes:} An obvious application of Kolmogorov's extension theorem is the construction of basic stochastic processes like Markov chains \cite[Section~11]{Kallenberg2020}, the Poisson process \cite[Section~13]{Kallenberg2020} and Brownian motion \cite[Section~14]{Kallenberg2020}. (Note that other Gaussian processes -- indexed by $[0, \infty)$ or $\mathbb R$ -- might as well be constructed the same way. In addition, fields like the Gaussien free field indexed by $\mathbb R^d$ \cite{sheffield2007gaussian} can also be given.) For these tasks, we would have to define the finite-dimensional distributions (Poisson and normally distributed, respectively), and apply the extension theorem. This task requires the formalization of multi-dimensional Poisson and normal distributions, which -- in textbooks -- is usually done using characteristic functions. Since these are not yet part of {\tt mathlib}, we postpone this task to the future.

@@

\noindent
{\em Theorem of Ionescu-Tulcea and related results:} While Kolmogorov's extension Theorem gives a result for arbitrary $\iota$, but requires some properties of the family $(\alpha_t)_{t\in\iota}$, the Theorem of Ionescu-Tulcea \cite[Theorem 8.24]{Kallenberg2020} can only deal with countable $\iota$, but has no restrictions on $(\alpha_t)_{t\in\iota}$. The proof of the latter, however, uses transition kernels (and an inductive construction of a content, which can be extended to a measure) rather than projective families. However, continuity of a measure at $\emptyset$ is used as well and a formalization would complement our work. In addition, using the Theorem of Ionescu-Tulcea one can define infinitely many independent random variables (arbitrary $\iota$) on a joint probability space \cite[Theorem 8.24]{Kallenberg2020} without requirements on $(\alpha_t)_{t\in\iota}$. All of these results can be used in order to construct countably many independent random variables, which are frequently used in concrete constructions in probability theory. Examples are (one direction of) the Borel-Cantelli-Lemma \cite[Theorem 4.28]{Kallenberg2020}, Kolmogorov's 0-1-law \cite[Theorem 4.13]{Kallenberg2020}, random walks \cite[Chapter 12]{Kallenberg2020}, branching processes  \cite[Chapter 13]{Kallenberg2020}, and percolation  \cite[Chapter 2.4]{Klenke2013}, to name just a few. While the first two are already formalized in \leanline{mathlib}, it is only clear due to the Kolmorov extension theorem (or the Theorem of Ionesu-Tulcea) that infinitely many independent random variables can indeed be defined on a joint probability space. In this sense, our contribution is also important for giving more sense to already formalised parts of \leanline{mathlib}.

\section{Formalization of the Daniell-Kolmogorov extension}
We start by stating the exact result. We are going to formulate the main result in a modern fashion, as e.g.\ found in Theorem 2.2 of \cite{rao1971projective}, Theorem 7.7.1 of Volume~2 of \cite{bogachev2007measure}, Theorem 15.26 of \cite{guide2006infinite}, or \cite{border1998expository}. Note that these formulations split general assumptions on the underlying space(s) (e.g.\ a metric property) from the property which is needed in the proof (inner regularity with respect to compact sets). Other -- highly readable -- references such as \cite{Billingsley1995} state the extension theorem only in special cases such as $\alpha_t = \mathbb R$ for all $t$. 
%In addition, there are still new proofs found for the general extension theorem \cite{chin2019new}, but they require some results not yet formalized in mathlib (Borel isomorphism of $\mathbb R$ to $\{0,1\}^{\mathbb N}$), so we do not follow this route.

\subsection{Formulating the result}
Before we can state the result, we begin with some mathematical notions. We start off from metric spaces, but quickly introduce projective families of measures. In all definitions, $\alpha$ will be some set (or type, since {\tt Lean} is a dependently typed language, \leanline{α : Type _}).

@@

\begin{remark}[Metric and topological spaces in {\tt mathlib}] \mbox{}
  \begin{enumerate}
    \item \sloppy In \leanline{Lean}, we have the \leanline{class
      PseudoEMetricSpace (α : Type u) extends EDist}, where
      \leanline{EDist} comes with an extended distance \leanline{edist
        : α → α → ENNReal}, and the extension to
      \leanline{PseudoEMetricSpace} consists of the components,
      \leanline{edist_self}, \leanline{edist_comm},
      \leanline{edist_triangle} providing the properties of the
      pseudo-metric, and \leanline{toUniformSpace},
      \leanline{uniformity_edist} defining a uniform space from the
      extended pseudo-metric. Such a space does not come with a
      metric, but with a filter on $\alpha \times \alpha$, which
      describes which points in $\alpha$ are near. For example, the
      diagonal of $\alpha \times \alpha$ is a subset of all sets in
      the uniformity; see \cite{james2013topologies} for details. We
      note that a uniform space with a countably generated uniformity
      filter is pseudometrizable, i.e.\ there exists a
      pseudo-metric-space structure that generates the same
      uniformity; see \leanline{UniformSpace.pseudoMetrizableSpace},
      which formalizes a result stated in
      \cite{melikhov2011metrizable}.
    \item \sloppy For topological spaces, we have \leanline{class
      TopologicalSpace(α : Type u) : Type u}, which comes with
      \leanline{IsOpen : Prop}, \leanline{isOpen_univ : IsOpen
        Set.univ}, \leanline{isOpen_inter} and
      \leanline{isOpen_sUnion}, which are exactly the properties of a
      topological space described above.
    \item Any (extended pseudo-)metric on $\alpha$ defines a topology,
      namely the topology generated by $\mathcal H := \{\{y:
      r(x,y)<\varepsilon\}: x \in \alpha, \varepsilon \in
      (0,\infty)\}$. Actually, in \leanline{mathlib}, as we have seen
      above, every extended pseudo-metric defines a uniform space, and
      a uniform space is a \leanline{class UniformSpace(α : Type u)
        extends TopologicalSpace , UniformSpace.Core} (where the
      uniformity is defined in \leanline{UniformSpace.Core}), which
      connects the \leanline{uniformity} with the topology using
      \leanline{∀ (s : Set α), IsOpen s ↔ ∀ (x : α), x ∈ s → {p |
          p.fst = x → p.snd ∈ s} ∈ uniformity}. In particular, a
      uniform space defines a topological space, which can be used by
      the typeclass system.
  \end{enumerate}
\end{remark}

@@

\begin{remark}[Measur(abl)e spaces in \leanline{mathlib}]
  \begin{enumerate}
  \item \sloppy The \leanline{class MeasurableSpace α} is very similar
    to a topological space in \leanline{mathlib}, since it comes with
    \leanline{MeasurableSet' : Set α → Prop},
    \leanline{measurableSet_empty : MeasurableSet' ∅},
    \leanline{measurableSet_compl : (s : Set α) → MeasurableSet' s →
      MeasurableSet' sᶜ}, and \leanline{measurableSet_iUnion},
    i.e.\ properties (i)-(iii) from Definition~\ref{def:sigma}.1.
  \item We are using outer measures in our construction. In
    \leanline{mathlib}, this is \leanline{structure
      MeasureTheory.OuterMeasure α} which is a function
    \leanline{measureOf : Set α → ℝ≥0∞} such that \leanline{measureOf
      ∅ = 0} (i.e.\ the empty set has measure~$0$), \leanline{∀ {s₁ s₂
        : Set α}, s₁ ⊆ s₂ → measureOf s₁ ≤ measureOf s₂}
    (i.e.\ monotonicity) and \leanline{∀ (s : ℕ → Set α), measureOf (⋃
      (i : ℕ), s i) ≤ ∑' (i : ℕ), measureOf (s i)} (which we call
    $\sigma$-subadditivity). \\ Note that an outer measure is defined
    on $2^\alpha$, whereas a measure is only defined on a subset (the
    $\sigma$-algebra of measurable sets). More precisely: A
    \leanline{MeasureSpace α} extends a \leanline{MeasurableSpace α}
    by a \leanline{Measure α} (which is usually called
    \leanline{volume}). For the measure, we have: \leanline{structure
      MeasureTheory.Measure α [MeasurableSpace α] extends
      MeasureTheory.OuterMeasure}. Here $\sigma$-additivity of a
    \leanline{Measure α} reads
\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  m_iUnion : ∀ ⦃f : ℕ → Set α⦄,
    (∀ (i : ℕ), MeasurableSet (f i)) →
    Pairwise (Disjoint on f) →
    ↑toOuterMeasure (⋃ (i : ℕ), f i) =
    ∑' (i : ℕ), ↑toOuterMeasure (f i)
\end{minted}    
  \end{enumerate}
\end{remark}

@@

\begin{remark}[Generated $\sigma$-algebra, image measure]
  We will frequently need two basic results: 
  \begin{enumerate}
  \item The intersection of any number of $\sigma$-algebras is again a
    $\sigma$-algebra. For this reason, if $\mathcal H \subseteq
    2^\alpha$, we define the $\sigma$-algebra $\sigma(\mathcal H) :=
    \bigcap_{\mathcal F \supseteq \mathcal H \text{ $\sigma$-algebra}}
    \mathcal F$. This is called the $\sigma$-algebra generated from
    $\mathcal H$; see \leanline{MeasurableSpace.generateFrom}.  \\ In
    particular, if $\mathcal O$ is a topology on $\alpha$, we call
    $\mathcal B := \sigma(\mathcal O)$ the Borel $\sigma$-algebra on
    $\alpha$. In \leanline{mathlib}, this is
\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  def borel (α : Type u) [TopologicalSpace α] :
    MeasurableSpace α :=
    generateFrom { s : Set α | IsOpen s }
\end{minted}
    In addition, \leanline{mathlib} provides a similar notion, which is
\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  class OpensMeasurableSpace (α : Type*)
    [TopologicalSpace α] [h : MeasurableSpace α] :
    Prop where borel_le : borel α ≤ h
\end{minted}
    Here, all open sets are measurable, so the $\sigma$-algebra defining
    \leanline{h} might be larger than the Borel $\sigma$-algebra. We will
    need both notions in our proofs.
  \item Let $(\alpha, \mathcal F)$ and $(\beta, \mathcal G)$ be
    measurable spaces. Some $f : \alpha \to \beta$ is called
    measurable (with respect to $\mathcal F$ and $\mathcal G$) if
    $f^{-1} \mathcal G \subseteq \mathcal F$. In \leanline{mathlib},
    see \leanline{Measurable} and note that this notion usually comes
    with some \leanline{m : Measure α} and $f$ is almost everywhere
    measurable (\leanline{AEMeasurable}) if it coincices with some
    \leanline{Measurable g}, \leanline{m}-almost everywhere.\\ In this
    case, if $\mu$ is a measure on $\mathcal F$, the measure $\nu:
    \mathcal G \to [0, \infty], \nu(B) := \mu (f^{-1}(B))$ is called
    the image (or push-forward) measure of $\mu$ under $f$. We write
    $\nu := f_\ast \mu$; see \leanline{MeasureTheory.Measure.map}.
  \end{enumerate}
\end{remark}

We now come to our contribution, which we will PR to mathlib: The
extension theorem is on finite measures on product spaces. In
particular, the product is taken over a set of arbitrary
cardinality. The next definition covers the important concept of a
projective family of measures. In short term, we define measures on
any finite subset of indices in a consistent way. Recall that finite
sets (of some type \leanline{ι}) are formalized with \leanline{Finset
  ι}. Such a \leanline{Finset} is defined as a \leanline{Multiset} (a
quotient type of the \leanline{List}-type with respect to permutations
of the list) together with a proof that it contains no duplicates.

@@

\noindent
With this, we can define the projetive family as follows. Note that
\leanline{Lean} automatically uses the subset relation as
\leanline{[Preorder (Finset ι)]} when \leanline{isProjective} is
called.

@@

Let us remark that (i) \leanline{Lean} indicates a coercion by
\leanline{↑}, which in this case is from \leanline{Finset} to
\leanline{Set} and (ii) we have \leanline{PairwiseDisjoint (s : Set ι)
  (f : ι → α)} iff the images of any distinct elements of \leanline{ι}
under \leanline{f} are different. (Hence, if \leanline{f = id}, the
usual definition of pairwise disjoint sets unfolds.)

@@

\noindent
For additive contents, we need two definitions. \leanline{AddContent}
defines a set-function \leanline{tuFun} on \leanline{Set α}, whereas
\leanline{extendContent} comes with \leanline{m : ∀ s : Set α, s ∈ C
  → ℝ≥0∞}.

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  structure AddContent (C : Set (Set α)) where
    toFun : Set α → ℝ≥0∞
    empty' : toFun ∅ = 0
    add' : ∀ (I : Finset (Set α))
      (_h_ss : ↑I ⊆ C) (_h_dis : PairwiseDisjoint
        (I : Set (Set α)) id)
      (_h_mem : ⋃₀ ↑I ∈ C),
      toFun (⋃₀ I) = ∑ u in I, toFun u

  noncomputable def extendContent
    (hC : SetSemiring C)
    (m : ∀ s : Set α, s ∈ C → ℝ≥0∞)
    (m_empty : m ∅ hC.empty_mem = 0)
    (m_add :
      ∀ (I : Finset (Set α)) (h_ss : ↑I ⊆ C)
      (_h_dis : PairwiseDisjoint (I : Set (Set α)) id)
      (h_mem : ⋃₀ ↑I ∈ C),
      m (⋃₀ I) h_mem = ∑ u : I, m u (h_ss u.prop)) :
    AddContent C 
\end{minted}

\noindent
In these formalizations, we introduced $m(\emptyset)=0$ as a separate
hypothesis with \leanline{empty'}. Note that this also follows
formally from $m(\emptyset) = m(\emptyset \cup \emptyset) = 2
m(\emptyset)$. However, since \leanline{Finset}s are assumed to have
distinct elements, we cannot have \leanline{∅} twice in \leanline{I}.

For the concrete application we have in mind, we need to be more
specific which semiring to use, based on a projective family of
probability measures. We do this using \leanline{kolContent}, which is
based on \leanline{kolmogorovFun}, which takes a family of measures
and a cylinder set as input:

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  noncomputable def kolmogorovFun
    (P : ∀ J : Finset ι, Measure (∀ j : J, α j))
    (s : Set (∀ i, α i))
    (hs : s ∈ cylinders α) : ℝ≥0∞ :=
    P (cylinders.finset hs) (cylinders.set hs)

@@

Here, \leanline{kolmogorovFun_empty} and
\leanline{kolmogorovFun_additive} give proofs for \leanline{m_empty}
and \leanline{m_add} in \leanline{extendContent} as applied to a
\leanline{kolmogorovFun}.

\noindent
From the next lemma, we will need monotonicity and sub-additivity, as
well as $\sigma$-additive $\Rightarrow$ $\sigma$-sub-additive in the
proof of the Carathéodory extension, Theorem~\ref{T:masseind}. The
formalization of the next result comes in several lemmas, which we do
not show here.

@@

The above definition closely resembles its formalization in
\leanline{mathlib}:
